---
title: "Visual Analytics Exercise 1-5"
output:
  html_document:
    df_print: paged
---


## Visual Analytics for Policy and Management: Exercise 1

## Univariate Case

## By Kimberly Pearson 

Data to be used: 
```{r getdata, eval=TRUE}
link='https://github.com/EvansDataScience/VisualAnalytics_2_tabularData/raw/master/data/eduwa.rda'

#getting the data TABLE from the file in the cloud:
load(file=url(link))
```

## Exercise 1: Create a GGPlot Barchart.

```{r tableLocaleProp, eval=TRUE}
absoluteT=table(eduwa$LocaleType,exclude = 'nothing')

ToPlot=prop.table(absoluteT)*100

names(ToPlot)[5]='Unknown'

titleText='Where are Public Schools located in WA in 2019?'
sourceText='Source: US Department of Education'

library(ggplot2)
ToPlotOrd=sort(ToPlot)

tableFreq=as.data.frame(ToPlotOrd)

names(tableFreq)=c("locale","pct")
```

```{r ReorderTable, eval=TRUE}
library(ggplot2)

base= ggplot(data = tableFreq, 
             aes(x = locale,
                 y = pct)) 
plot1 = base + geom_bar(fill ="gray",
                        stat = 'identity') # y is just what it is!

plot2 = plot1 + labs(title=titleText,
                     x =NULL, 
                     y = NULL,
                     caption = sourceText)

plot3 = plot2 + geom_hline(yintercept = 25, #where
                           linetype="dashed", 
                           size=1.5, #thickness
                           alpha=0.5) #transparency

library(scales)

# customize Y axis
plot4 = plot3 + scale_y_continuous(breaks=c(0,10, 25,40,50),
                                 limits = c(0, 50), # expand = c(0, 0),
                                 labels=scales::unit_format(suffix = '%')) 

plot5 = plot4 + theme(panel.background = element_rect(fill = "white",
                                                    colour = "grey50"),
                    plot.caption = element_text(hjust = 0), # default was 1
                    plot.title = element_text(hjust = 0.5))
plot6 = plot5 + geom_text(aes(y = pct ,
                            label = paste0(round(pct,2), '%')),
                        vjust=1, # if flipping 'hjust'
                        size = 3)

plot6 + coord_flip()
```

## Exercise 2: Add the elements missing in the last Lollipop plot. 

```{r lolliplot, eval=TRUE}
tableFreqO=tableFreq[order(tableFreq$pct),]

tableFreqO$gap=tableFreqO$pct-25

tableFreqO$PositiveGap=ifelse(tableFreqO$gap>0,T,F)

base = ggplot(tableFreqO, aes(locale,gap,color=PositiveGap,
                              label = round(gap,3))) #  change
lolliplot1=base + geom_segment(aes(y = 0, 
                                   x = locale, 
                                   yend = gap, 
                                   xend = locale), color = "gray") 
lolliplot2=lolliplot1 + geom_point() 

lolliplot3= lolliplot2 + scale_x_discrete(limits=tableFreqO$locale) 
# annotating and moving the text on the horizontal
lolliplot4 = lolliplot3 + geom_text(nudge_x=0.3,show.legend = FALSE) 

lolliplot5 = lolliplot4 + labs(title=titleText,
                     x =NULL, 
                     y = NULL,
                     caption = sourceText)

lolliplot6 = lolliplot5 + theme(
                    plot.caption = element_text(hjust = 1), # default was 1
                    plot.title = element_text(hjust = 0.5))

lolliplot6
```

## Exercise 3: Create a Box Plot for the Highest Grades Offered. 

```{r titletext2, eval=TRUE}
titleText2='Distribution of Highest Grades Offered at WA Public Schools'
```

```{r ggbox, eval=TRUE}
library(ggplot2)
box1 = ggplot(eduwa, aes(y=as.numeric(High.Grade))) +   theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
box1 = box1 + geom_boxplot() + coord_flip() # to show it horizontally
# Create labels that use the original ordinal levels
ordLabels= levels(eduwa$High.Grade)
# Add the labels (based on the ordLabels) to the plot 
# Breaks are from 1 to 15 because there are 15 grades
box2 = box1 + scale_y_continuous(labels=ordLabels,breaks=1:15)
box3 = box2 + labs(title=titleText2,
                     x =NULL, 
                     y = NULL,
                     caption = sourceText)
box4 = box3 + ylab("Grades Offered") 
box5 = box4 + theme(plot.title = element_text(family="Times", color="black", size=14, face="bold", hjust = 0.5), plot.caption = element_text(family="Times", color="black", size=10, face="plain", vjust = 0), axis.title.x = element_text(family="Times", color="black", size=10, face="italic", vjust = 0.5, hjust = 0.5))
box5
```

## Exercise 4: Make a histogram for the variable FREE LUNCH, and make sure it has all the right elements, and get rid of unnecessary elements. 

```{r}
#ggplot
eduwa2=eduwa[complete.cases(eduwa$Free.Lunch),]
library(ggplot2)
base= ggplot(eduwa2,aes(x = Free.Lunch))  
h1= base + geom_histogram(breaks=seq(0, 1800, 100), 
                          boundary=0, 
                          col="red", 
                          fill="blue", 
                          alpha=.3, 
                          na.rm = TRUE, 
                          inherit.aes = TRUE)

h1= h1 + stat_bin(binwidth = 100, aes(label=..count..), cex=2.5, 
                  geom = "text", boundary = 0,vjust=-0.5)

h1= h1 + labs(title="Rates at Which Schools Offer Free Lunch to Students", subtitle =sourceText, x="Quantity of Free Lunches", y="Number of Schools") + 
  ylim(c(0,900))
h2 = h1 + scale_x_continuous("Quantity of Free Lunches", breaks=seq(0, 1800, 100), limits = c(0,1800))
h2
```

## Exercise 5: Make a Histogram and a Density Plot using GGplot for the mean values of Student.Teacher.Ratio

```{r}
eduwa3=eduwa[complete.cases(eduwa$Student.Teacher.Ratio),]

tapply(eduwa3$Student.Teacher.Ratio, 
       eduwa3$County, 
       mean,
       na.rm=T)

meanValues=tapply(eduwa3$Student.Teacher.Ratio, 
                  eduwa3$County, 
                  mean)

df.T=as.data.frame(meanValues)
names(df.T)=c('StudentTeacherRatio')
```

```{r STRhist, eval=TRUE}
library(ggplot2)
ggplot(data = df.T, aes(x= StudentTeacherRatio)) +
  geom_histogram(breaks=seq(8, 24, by=2),
                       col="red",
                       fill="blue",
                       alpha=.2,
                       na.rm = TRUE,
                       inherit.aes = TRUE) + 
  labs(title="WA Student Teacher Ratios", subtitle =sourceText, x="Student Teacher Ratios", y="Number of Schools") + 
  ylim(c(0,20)) +
  stat_bin(binwidth = 2, aes(label=..count..), cex=2.5, 
                  geom = "text", boundary = 0,vjust=-0.5) +
  scale_x_continuous(breaks=seq(8, 24, 2), limits = c(8,24)) +
  geom_vline(aes(xintercept = mean(df.T$StudentTeacherRatio),colour="mean"),
                           show.legend  = TRUE) +
  geom_vline(aes(xintercept = median(df.T$StudentTeacherRatio),colour="median"),
                           show.legend  = TRUE) + 
                scale_color_manual(name = NULL, 
                                   values = c(median = "blue", mean = "green"))
```

```{r STRDensity, eval=TRUE}
base= ggplot(data=df.T, aes(x=StudentTeacherRatio)) + 
  geom_density(col="red",
               fill="blue",
               alpha=.2) + 
  labs(title="Density Plot of Average WA Student Teacher Ratios", subtitle =sourceText, x="Student Teacher Ratios", y="Number of Schools") + 
  ylim(c(0,.18))

density1 = base + geom_vline(aes(xintercept = mean(eduwa3$Student.Teacher.Ratio),colour="mean"),
                           show.legend  = TRUE) +
                geom_vline(aes(xintercept = median(eduwa3$Student.Teacher.Ratio),colour="median"),
                           show.legend  = TRUE) + 
                scale_color_manual(name = NULL, 
                                   values = c(median = "blue", mean = "red"))

density2 = density1 +  scale_x_continuous(breaks=seq(0,24,2))
density2 
```

